Parameter Inference for a Simple Stochastic Hydrological Model

### 2015-07-16 TeXFiles/Manuscript_Albert_Ulzega.tex
tex files updated to NewJPhys template

\documentclass[11pt, a4paper]{article}

\usepackage{a4wide}
\usepackage{amsmath,amssymb,amsthm,amscd}
\usepackage{natbib}
\usepackage[pdftex]{graphicx}
\usepackage[affil-it]{authblk}
\usepackage{sidecap}
\usepackage[font=small,labelfont=bf]{caption}

%\usepackage{showkeys}

\usepackage[all]{xy}

\newcommand{\bb}[1]{\begin{equation}\label{#1}}
\newcommand{\ee}{\end{equation}}
\newcommand{\vc}[1]{{\bf #1 }}
\renewcommand{\vec}[1]{{\bf #1}}
\newcommand{\Sum}[1]{\sum\limits_{#1}}
\newcommand{\Int}[1]{\int\limits_{#1}}
\newcommand{\Prod}[1]{\prod\limits_{#1}}
\newcommand{\const}{\operatorname{const}}
\newcommand{\ul}{\underline}
\newcommand{\ol}{\overline}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\bbA}{\mathbb A}
\newcommand{\bbR}{\mathbb R}
\newcommand{\bbN}{\mathbb N}
\newcommand{\bbZ}{\mathbb Z}
\newcommand{\bbC}{\mathbb C}
\newcommand{\bbT}{\mathbb T}
\newcommand{\QB}{\mathcal H}
\newcommand{\PQB}{\tilde{\mathcal H}}
\newcommand{\CQB}{\mathcal H'}
\newcommand{\Ad}{\operatorname{Ad}}
\newcommand{\ad}{\operatorname{ad}}
\newcommand{\gh}{\operatorname{gh}}
\newcommand{\Der}{\operatorname{Der}}
\newcommand{\Aut}{\operatorname{Aut}}
\newcommand{\aut}{\operatorname{aut}}
\newcommand{\Diff}{\operatorname{Diff}}
\renewcommand{\dim}{\operatorname{dim}}
\newcommand{\Vol}{\operatorname{Vol}}
\newcommand{\Pf}{\operatorname{Pf}}
\newcommand{\Cov}{\operatorname{Cov}}

\newcommand{\g}{{\bf g}}
\newcommand{\h}{{\bf h}}
\renewcommand{\d}{{\bf d}}
\newcommand{\e}{{\bf e}}
\renewcommand{\t}{{\bf t}}
\newcommand{\lpartial}{\overset{\rightarrow}{\partial}}
\newcommand{\rpartial}{\overset{\leftarrow}{\partial}}
\newcommand{\lder}[1]{\overset{\rightarrow}{\vc #1}}
\newcommand{\rder}[1]{\overset{\leftarrow}{\vc #1}}
\newcommand{\supp}{\operatorname{supp}}
\newcommand{\im}{\operatorname{Im}}
\renewcommand{\ker}{\operatorname{Ker}}
\newcommand{\Var}{\operatorname{Var}}

%define the \identy operator
\def\litID{{\sf id}}
\def\identy{{\mathsurround0pt\mathchoice{\textidenty}{\textidenty}{\scptidenty}{\scptidenty}}}
\def\scptidenty{\setbox0\hbox{$\scriptstyle1$}\bothidenty}
\def\textidenty{\setbox0\hbox{$1$}\bothidenty}
\def\bothidenty{\rlap{\hbox to.97\wd0{\hss\vrule height.06\ht0 width.82\wd0}}
 \copy0\rlap{\kern-.36\wd0\vrule height1.05\ht0 width.05\ht0}\kern.14\wd0}
%end of definition \identy

\newcommand{\bbM}{\mathbb M}
\newcommand{\bt}{\pmb\theta}
\newcommand{\Bt}{\pmb\Theta}
\newcommand{\dt}{\mathit{dt}}
\newcommand{\eq}{\text{eq}}

\begin{document}
\title{Parameter Inference for a Simple Stochastic Hydrological Model}

\author{C. Albert\thanks{\noindent Email: carlo.albert@eawag.ch} \,,
S. Ulzega}
\affil{Eawag, Swiss Federal Institute of Aquatic Science and Technology, 8600 D\"ubendorf, Switzerland}
\maketitle

In this paper, we study the feasibility of a full Bayesian inference, for simple conceptual rainfall-runoff models with scale-invariant noise.
We start with a single linear reservoir and describe the dynamics of its water content, $S(t)$, by means of the stochastic differential equation
\begin{equation}\label{sde}
\dot{S}(t) = r(t) - \frac{1}{K}\left(1+\frac{\gamma}{2}\right) S(t)
+
\sqrt{\frac{\gamma}{K}} S(t){\eta}(t)\,,
\end{equation}
where $r(t)$ describes the rain input and $\eta(t)$ denotes white noise, i.e.,
\begin{equation}\label{whitenoise}
\langle\eta(t)\eta(t')\rangle = \delta(t-t')\,.
\end{equation}
Due to the state-dependence of the noise term, eq. (\ref{sde}) is ill-defined unless we specify the discretization convention.
For computational reasons, we work with the {\em Stratonovich} convention.
Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{eq}=Kr_0$ (see below).

The scale-invariance of the noise term renders the noise parameter $\gamma$ dimensionless.
Before we proceed, we replace the state variable $S(t)$ and parameter $K$ by dimensionless quantities $q(t)$ and $\beta$, respectively, by means of the transformations
\begin{equation}
  \beta=\sqrt{\frac{T\gamma}{K}}\,,\quad
  S(t)=\frac{T\gamma r(t)}{\beta^2}e^{\beta q(t)}\,.
\end{equation}
W.r.t. these new  variables and parameters, model equation (\ref{sde}) becomes
\begin{equation}
  \dot q(t)
  =
  \frac{\beta}{T\gamma}e^{-\beta q(t)}
  -
  \frac{1}{T}\rho(t)
  +
  \frac{1}{\sqrt{T}}\eta(t)\,,
\end{equation}
with
\begin{equation}
  \rho(t)
  =
  \frac{T}{\beta}\frac{d}{dt}\ln r(t)
  +
  \frac{(2+\gamma)\beta}{2\gamma}\,.
\end{equation}

%Given rainfall and runoff time-series, we want to infer the two system parameters $K$ and $\gamma$, measuring, respectively, the retention time and the noise.

According to \cite{lau_2007_StateDepDiff}, the probability, $P(q_1,T|q_0,0)$, of finding the system in a state $q$ at time $t = T$ given that it was in an initial state $q_0$ at time $t = 0$, can be expressed in the form of a path integral as
\begin{equation}\label{pathint}
P(q_1,T|q_0,0)
=
\frac{1}{Z}
\int
e^{-{\mathcal S}[q,\dot q]}
\delta(q(T)-q_1)
\delta(q(0)-q_0)
\mathcal{D}q \,,
\end{equation}
where the integral extends over all paths $q:[0,T]\rightarrow \mathbb R$.
The path-measure $\mathcal Dq$ is formally written as an infinite product
\begin{equation}\label{pathmeasure_q}
{\mathcal Dq}
=
\prod_{t}
dq(t)\,.
\end{equation}
The {\em action} is a functional on the space of paths and reads as
%\begin{equation}\label{action}
%  \tilde{\mathcal S}[\dot{S}(t),S(t)]
%  =
%  \int_0^T
%  dt
%  \left\{
%    \frac{K}{2\gamma S^2(t)}
%    \left(
%        \dot S(t)
%        -
%        r(t)
%        +
%        \frac{1+\gamma}{K}S(t)
%    \right)^2
%    -
%    \frac{2+\gamma}{4K}
%  \right\}
% \,.
%\end{equation}
\begin{equation}\label{action1}
{\mathcal S}[{q},\dot q]
=
\frac{1}{T}
\int_0^T \dt \left\{
\frac{1}{2}
\left(
    T\dot q(t)
    +
    \rho(t)
    -
    \frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2
    -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
\right\} \,.
\end{equation}
Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $q(t)$.

We introduce the time-dependent {\em Hamiltonian}
\begin{equation}\label{H}
  \mathcal{H}(q,t)
  =
  \frac{1}{\gamma}e^{-\beta q}+q\rho(t)\,,
\end{equation}
and rewrite action (\ref{action1}) as,
\begin{multline}\label{action}
{\mathcal S}[{q},\dot q]
=
\frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    T\frac{\partial \mathcal{H}(q,t)}{\partial t} -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
     %\frac{(2+\gamma)\beta^2}{4\gamma}
\right\}
\\
+ \mathcal{H}(q(T),T) - \mathcal{H}(q(0),0)
\\
= \frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    Tq(t)\dot\rho(t) -
     \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
%    \frac{(2+\gamma)\beta^2}{4\gamma}
\right\}
\\
+
    \frac{1}{\gamma}e^{-\beta q(T)}+q(T)\rho(T)
   -\frac{1}{\gamma}e^{-\beta q(0)}-q(0)\rho(0)
\,.
\end{multline}

Next, we calculate the equilibrium distribution $P_{\text{eq}}(q) = \lim_{T\rightarrow\infty} P(q,T|q_0,0)$ in the simple case of a constant rain input $r(t)\equiv r_{0}$. Systems at thermal equilibrium must fulfill detailed balance, i.e.,
\begin{equation}\label{detailed_balance}
P(q_1 t_1 | q_0 t_0 ) P_{\text{eq}}(q_0) = P(q_0 t_1 | q_1 t_0 ) P_{\text{eq}}(q_1) \,.
\end{equation}
After plugging in (\ref{pathint}) and (\ref{action}), and using the transformation $q(t) \rightarrow q(-t)$, which maps paths from $q_0$ to $q_1$ to paths from $q_1$ to $q_0$, we get, since $\dot\rho(t)= 0$,
\begin{equation}\label{Peq}
  P_{\text{eq}}(q)
  \propto
  e^{-2\mathcal{H}(q)}\,.
\end{equation}
Transforming back to the original variables, it turns out that $P_{\text{eq}}(S)$ is described by an inverse gamma distribution with scale parameter $2Kr_{0}/\gamma$ and shape parameter $(2+\gamma)/\gamma$, i.e.,
\begin{equation}\label{inverse_gamma}
  P_{\text{eq}}(S)
  \propto
  S^{-2(1+\gamma)/\gamma}e^{-2Kr_{0}/(\gamma S)}\,,
\end{equation}
whose mean equals the equilibrium solution of the unperturbed system ($\gamma=0$),
\begin{equation}
  \langle S\rangle_{\eq}=Kr_{0},
\end{equation}
and whose variance, for $\gamma< 2$, is given by
\begin{equation}
  \langle (S - \langle S\rangle_{\eq})^2\rangle_{\eq}
  =
  K^2r_{0}^2
  \frac{\gamma}{2-\gamma}\,.
\end{equation}
Note that the variance diverges, for $\gamma\geq 2$.
The power-law decay of the inverse gamma distribution is reminiscent of the scale-invariance of the error model.

@@

%The {\em likelihood} function for the data (\ref{data}) is, up to a factor that doesn't depend on the model parameters, given by the path-integral
%\begin{equation}\label{likelihood}
%  f(\vc y | \bt)
%  \propto
%  \int
%  \exp\bigg[
%    -{\tilde{\mathcal S}}[\dot{S},S]
%    -\frac{1}{2}
%    \sum_{i=1}^{n+1}
%    \frac
%    {(y_i-S(t_i)/K)^2}
%    {\sigma^2}
%  \bigg]
%  {\mathcal D} S(\bt)\,,
%\end{equation}
%with the parameter vector $\bt=(K,\gamma)^T$.
%Choosing for the parameters $K$ and $\gamma$ the improper prior given by
%
%\begin{equation}\label{prior}
%  f(\bt)
%  =  f(K) f(\gamma) = \frac{1}{K \gamma} \,,
%\end{equation}
%
%{\em parameter inference} means generating a sufficiently large sample of parameter vectors from the posterior
%\begin{equation}\label{posterior}
%  f(\bt|\vc y)\propto f(\vc y|\bt)f(\bt)\,.
%\end{equation}

The goal is to infer the posterior parameter distribution, for the parameters $\bt=(\beta,\gamma)^T$, based on the measured time series $\vc r$ and $\vc y$.
For the time being we assume constant priors.
Then, the posterior reads as
\begin{equation}\label{posterior_pathint}
  f(\bt | \vc y, \vc r)
  \propto
  \int
  \exp\bigg[
    -{{\mathcal S}}[{q},\dot q]
    -\frac{1}{2}
    \sum_{s=1}^{n+1}
    \frac
    {(\ln(y_s/r_s)-{\beta q_s})^2}
    {\sigma^2}
    %-\ln(K\gamma)
  \bigg]
  {\mathcal Dq}
  \,.
\end{equation}
Since the integral on the r.h.s. of (\ref{posterior_pathint}) is very expensive to calculate, we will sample, simultaneously, parameter vectors $\bt$ and discretized system realizations $q(t)$ directly from an appropriate discretization of the kernel of (\ref{posterior_pathint}).
Since the associated distribution is very high dimensional and strongly correlated, we employ the method called {\em Hamiltonian Monte Carlo} (HMC) introduced by \cite{duane_1987_HMC}.
This method interprets the exponent of the kernel of the posterior as the potential of a one-dimensional statistical mechanical system. Each degree of freedom, $q(t)$ and $\bt$ in our case, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$, respectively, and the system is defined by the  Hamiltonian
\begin{equation}\label{Hamiltonian}
    \mathcal{H}_{\text{HMC}}(q,\bt; p,{\pmb\pi})
    =
    K( p,{\pmb\pi}) + V( q,\bt)\,,
\end{equation}
with
\begin{equation}\label{K}
   K( p,{\pmb\pi})
   =
   \int_0^T \frac{ p^2(t)}{2m(t)}dt
   + \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,
\end{equation}
where $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
The posterior (\ref{posterior_pathint}) is expressed by the phase space path integral
\begin{equation}\label{phaseSpacePathInt}
    f(\bt | \vc y, \vc r)
  \propto
  \int
  e^{-\mathcal {H}_{\text{HMC}}(q,\bt; p,{\pmb\pi})}
  {\mathcal Dp}
  \wedge
  {\mathcal Dq}
  \wedge
  d{\pmb\pi}
  \,.
\end{equation}

The HMC method, which is a combination of the {\em Metropolis algorithm} by \cite{metropolis_1953_MRT2} and {\em molecular dynamics} methods \textbf{[ref?]}, iterates the following steps
\begin{enumerate}
  \item
  Momenta are sampled from the Gaussian distribution defined by (\ref{K}).
  \item
  A volume-preserving and time-reversible solution to a discretization of Hamilton's equations is calculated numerically.
  \item
  The discretization error on the energy preservation due to the previous step is corrected by a Metropolis acceptance/rejection step.
\end{enumerate}
The last step is the standard Metropolis algorithm, while the first two steps allow us to make relatively large jumps in phase space while maintaining a relatively large acceptance rate.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent
% ( open parenthesis    ) close paren.
% - hyphen              = equals sign
% | vertical bar        ~ tilde
% @ at sign             _ underscore
% { open curly brace    } close curly
% [ open square         ] close square bracket
% + plus sign           ; semi-colon
% * asterisk            : colon
% < open angle bracket  > close angle
% , comma               . full stop
% ? question mark       / forward slash
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ
% abcdefghijklmnopqrstuvwxyz
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt,a4paper,final]{iopart}
\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
%Uncomment next line if AMS fonts required
\usepackage{iopams}
\usepackage{graphicx}
\usepackage[breaklinks=true,colorlinks=true,linkcolor=blue,urlcolor=blue,citecolor=blue]{hyperref}
%\usepackage{amsmath,amssymb,amsthm,amscd}
%\usepackage{natbib}

\usepackage[all]{xy}
\newcommand{\eq}{\mbox{eq}}
\newcommand{\dt}{\mathit{dt}}
\newcommand{\bbM}{\mathbb M}
\newcommand{\bt}{\pmb\theta}
\newcommand{\Bt}{\pmb\Theta}
\newcommand{\vc}[1]{{\bf #1 }}

\begin{document}

\title[Parameter inference with stochastic models]{Parameter Inference for a Simple Stochastic Hydrological Model}

\author[cor1]{Carlo Albert}%$^{1,2}$
\address{Eawag, Swiss Federal Institute of Aquatic Science and Technology, 8600 Dübendorf, Switzerland}
%\address{$^2$Address Two, Neverland}
\ead{carlo.albert@eawag.ch}

\author{Simone Ulzega}
\address{Eawag, Swiss Federal Institute of Aquatic Science and Technology, 8600 Dübendorf, Switzerland}
\ead{simone.ulzega@eawag.ch}

%\author[cor1]{Author Three}
%\address{Address Four, Neverland}
%\eads{\mailto{author.three@mail.com}, \mailto{author.three@gmail.com}}

\begin{abstract}
\textbf{[TO BE WRITTEN]} This document describes the  preparation of an article using \LaTeXe\ and \verb"iopart.cls" (the IOP \LaTeXe\ preprint class file).
This class file is designed to help
authors produce preprints in a form suitable for submission to any of the
journals published by IOP Publishing.
Authors submitting to any IOP journal, i.e.\
both single- and double-column ones, should follow the guidelines set out here.
On acceptance, their TeX code will be converted to
the appropriate format for the journal concerned.

\end{abstract}

%Uncomment for PACS numbers title message
\pacs{00.00, 20.00, 42.10}
% Keywords required only for MST, PB, PMB, PM, JOA, JOB?
\vspace{2pc}
\noindent{\it Keywords}: Article preparation, IOP journals
% Uncomment for Submitted to journal title message
\submitto{\NJP}
% Comment out if separate title page not required
%\maketitle

\section{Introduction}

In this paper, we study the feasibility of a full Bayesian inference, for simple conceptual rainfall-runoff models with scale-invariant noise.
We start with a single linear reservoir and describe the dynamics of its water content, $S(t)$, by means of the stochastic differential equation
\begin{equation}\label{sde}
\dot{S}(t) = r(t) - \frac{1}{K}\left(1+\frac{\gamma}{2}\right) S(t)
+
\sqrt{\frac{\gamma}{K}} S(t){\eta}(t)\,,
\end{equation}
where $r(t)$ describes the rain input and $\eta(t)$ denotes white noise, i.e.,
\begin{equation}\label{whitenoise}
\langle\eta(t)\eta(t')\rangle = \delta(t-t')\,.
\end{equation}
Due to the state-dependence of the noise term, Eq.~(\ref{sde}) is ill-defined unless we specify the discretization convention.
For computational reasons, we work with the {\em Stratonovich} convention.
Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{\eq}=Kr_0$ (see below).

The scale-invariance of the noise term renders the noise parameter $\gamma$ dimensionless.
Before we proceed, we replace the state variable $S(t)$ and parameter $K$ by dimensionless quantities $q(t)$ and $\beta$, respectively, by means of the transformations
\begin{equation}
  \beta=\sqrt{\frac{T\gamma}{K}}\,,\quad
  S(t)=\frac{T\gamma r(t)}{\beta^2}e^{\beta q(t)}\,.
\end{equation}
W.r.t. these new  variables and parameters, the model equation (\ref{sde}) becomes
\begin{equation}
  \dot q(t)
  =
  \frac{\beta}{T\gamma}e^{-\beta q(t)}
  -
  \frac{1}{T}\rho(t)
  +
  \frac{1}{\sqrt{T}}\eta(t)\,,
\end{equation}
with
\begin{equation}
  \rho(t)
  =
  \frac{T}{\beta}\frac{d}{dt}\ln r(t)
  +
  \frac{(2+\gamma)\beta}{2\gamma}\,.
\end{equation}

According to Ref.~\cite{lau_2007}, the probability $P(q_1,T|q_0,0)$ of finding the system in a state $q$ at time $t = T$ given that it was in an initial state $q_0$ at time $t = 0$, can be expressed in the form of a path integral as
\begin{equation}\label{pathint}
P(q_1,T|q_0,0)
=
\frac{1}{Z}
\int
e^{-{\mathcal S}[q,\dot q]}
\delta(q(T)-q_1)
\delta(q(0)-q_0)
\mathcal{D}q \,,
\end{equation}
where the integral extends over all paths $q:[0,T]\rightarrow \mathbb R$.
The path-measure $\mathcal Dq$ is formally written as an infinite product
\begin{equation}\label{pathmeasure_q}
{\mathcal Dq}
=
\prod_{t}
dq(t)\,.
\end{equation}
The {\em action} is a functional on the space of paths and reads as
\begin{equation}\label{action1}
{\mathcal S}[{q},\dot q]
=
\frac{1}{T}
\int_0^T \dt \left\{
\frac{1}{2}
\left(
    T\dot q(t)
    +
    \rho(t)
    -
    \frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2
    -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
\right\} \,.
\end{equation}
Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $q(t)$.

We introduce the time-dependent {\em Hamiltonian}
\begin{equation}\label{H}
  \mathcal{H}(q,t)
  =
  \frac{1}{\gamma}e^{-\beta q}+q\rho(t)\,,
\end{equation}
and rewrite the action~(\ref{action1}) as,
\begin{eqnarray}\label{action}
{\mathcal S}[{q},\dot q] \nonumber
\\
= \frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    T\frac{\partial \mathcal{H}(q,t)}{\partial t} -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
     %\frac{(2+\gamma)\beta^2}{4\gamma}
\right\} \nonumber
\\
+ \mathcal{H}(q(T),T) - \mathcal{H}(q(0),0) \nonumber
\\
= \frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    Tq(t)\dot\rho(t) -
     \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
%    \frac{(2+\gamma)\beta^2}{4\gamma}
\right\}  \nonumber
\\
+
    \frac{1}{\gamma}e^{-\beta q(T)}+q(T)\rho(T)
   -\frac{1}{\gamma}e^{-\beta q(0)}-q(0)\rho(0)
\,.
\end{eqnarray}

Next, we calculate the equilibrium distribution $P_{\eq}(q) = \lim_{T\rightarrow\infty} P(q,T|q_0,0)$ in the simple case of a constant rain input $r(t)\equiv r_{0}$. Systems at thermal equilibrium must fulfill detailed balance, i.e.,
\begin{equation}\label{detailed_balance}
P(q_1 t_1 | q_0 t_0 ) P_{\eq}(q_0) = P(q_0 t_1 | q_1 t_0 ) P_{\eq}(q_1) \,.
\end{equation}
After plugging in (\ref{pathint}) and (\ref{action}), and using the transformation $q(t) \rightarrow q(-t)$, which maps paths from $q_0$ to $q_1$ to paths from $q_1$ to $q_0$, we get, since $\dot\rho(t)= 0$,
\begin{equation}\label{Peq}
  P_{\eq}(q)
  \propto
  e^{-2\mathcal{H}(q)}\,.
\end{equation}
Transforming back to the original variables, it turns out that $P_{\eq}(S)$ is described by an inverse gamma distribution with scale parameter $2Kr_{0}/\gamma$ and shape parameter $(2+\gamma)/\gamma$, i.e.,
\begin{equation}\label{inverse_gamma}
  P_{\eq}(S)
  \propto
  S^{-2(1+\gamma)/\gamma}e^{-2Kr_{0}/(\gamma S)}\,,
\end{equation}
whose mean equals the equilibrium solution of the unperturbed system ($\gamma=0$),
\begin{equation}
  \langle S\rangle_{\eq}=Kr_{0},
\end{equation}
and whose variance, for $\gamma< 2$, is given by
\begin{equation}
  \langle (S - \langle S\rangle_{\eq})^2\rangle_{\eq}
  =
  K^2r_{0}^2
  \frac{\gamma}{2-\gamma}\,.
\end{equation}
Note that the variance diverges, for $\gamma\geq 2$.
The power-law decay of the inverse gamma distribution is reminiscent of the scale-invariance of the error model.

@@

The goal is to infer the posterior parameter distribution, for the parameters $\bt=(\beta,\gamma)^T$, based on the measured time series $\vc r$ and $\vc y$.
For the time being we assume constant priors.
Then, the posterior reads as
\begin{equation}\label{posterior_pathint}
  f(\bt | \vc y, \vc r)
  \propto
  \int
  \exp\bigg[
    -{{\mathcal S}}[{q},\dot q]
    -\frac{1}{2}
    \sum_{s=1}^{n+1}
    \frac
    {(\ln(y_s/r_s)-{\beta q_s})^2}
    {\sigma^2}
    %-\ln(K\gamma)
  \bigg]
  {\mathcal Dq}
  \,.
\end{equation}
Since the integral on the r.h.s. of (\ref{posterior_pathint}) is very expensive to calculate, we will sample, simultaneously, parameter vectors $\bt$ and discretized system realizations $q(t)$ directly from an appropriate discretization of the kernel of (\ref{posterior_pathint}).
Since the associated distribution is very high dimensional and strongly correlated, we employ the method called {\em Hamiltonian Monte Carlo} (HMC) introduced by Ref.~\cite{duane_1987}.
This method interprets the exponent of the kernel of the posterior as the potential of a one-dimensional statistical mechanical system. Each degree of freedom, $q(t)$ and $\bt$ in our case, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$, respectively, and the system is defined by the  Hamiltonian
\begin{equation}\label{Hamiltonian}
    \mathcal{H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})
    =
    K( p,{\pmb\pi}) + V( q,\bt)\,,
\end{equation}
with
\begin{equation}\label{K}
   K( p,{\pmb\pi})
   =
   \int_0^T \frac{ p^2(t)}{2m(t)}dt
   + \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,
\end{equation}
where $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
The posterior (\ref{posterior_pathint}) is expressed by the phase space path integral
\begin{equation}\label{phaseSpacePathInt}
    f(\bt | \vc y, \vc r)
  \propto
  \int
  e^{-\mathcal {H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})}
  {\mathcal Dp}
  \wedge
  {\mathcal Dq}
  \wedge
  d{\pmb\pi}
  \,.
\end{equation}

The HMC method, which is a combination of the {\em Metropolis algorithm} of Ref.~\cite{metropolis_1953} and {\em molecular dynamics} methods \textbf{[ref?]}, iterates the following steps,
\begin{enumerate}
  \item
  Momenta are sampled from the Gaussian distribution defined by (\ref{K}).
  \item
  A volume-preserving and time-reversible solution to a discretization of Hamilton's equations is calculated numerically.
  \item
  The discretization error on the energy preservation due to the previous step is corrected by a Metropolis acceptance/rejection step.
\end{enumerate}
The last step is the standard Metropolis algorithm, while the first two steps allow us to make relatively large jumps in phase space while maintaining a relatively large acceptance rate.

Let us first write down the discretization of the path-integral (\ref{phaseSpacePathInt}).
Therefore, let us assume that the measurement time points $\left\{ y_s \right\}_{s=1,\dots, n+1}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$, with $t_1=0$ and $t_{n+1}=T$.
Each interval between two consecutive data points is further partitioned into $j$ bins, such that we have a total of $nj+1=N>>1$ discretization points.
Path-integral (\ref{posterior_pathint}) is approximated by an ordinary integral, with the approximate path-measure
\begin{equation}
  \mathcal Dp\wedge\mathcal Dq
  \approx
  \prod_i dp_i\wedge dq_i
\end{equation}
and the discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$
\begin{eqnarray}
   K( p,{\pmb\pi}) \approx
   \sum_{i=1}^N
   \frac{ p_i^2}{2m_q}\Delta t
   +
   \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,\label{Kdisc}
   \\
   V(q,\bt) \approx \frac{\Delta t}{T} \sum_{i=2}^{N}
   \left\{ \frac{1}{2} T^2 \dot q_i^2 + \frac{1}{2}
     \left( \rho_i-\frac{\beta}{\gamma}e^{-\beta q_i} \right)^2 -
    \frac{\beta^2}{2\gamma} e^{-\beta q_i} - T q_i\dot\rho_i \right\}  \nonumber
  \\
  +
  \frac{1}{\gamma}
  e^{-\beta q_N}
  +
  q_N \rho_{N}
  -
  \frac{1}{\gamma}
  e^{-\beta q_1}
  -
  q_1 \rho_{1}
  \nonumber
  \\
  +
  \sum_{s=1}^{n+1}
  \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta q_{(s-1)j+1}})^2}{2\sigma^2}
  \label{Vdisc}
   \,,
\end{eqnarray}
with
\begin{equation}
  \dot q_i = \frac{q_i-q_{i-1}}{\Delta t}\,,
  %\quad
%  \bar q_i = \frac{q_i+q_{i-1}}{2}\,,
\end{equation}
and
\begin{equation}\label{rhodisc}
\rho_i = \frac{T}{\beta} \frac{\ln(r_{i}/r_{i-1})}{\Delta t}
+
\frac{(2+\gamma)\beta}{2\gamma}
\,,\quad
\dot\rho_i = \frac{\rho_i-\rho_{i-1}}{\Delta t}\,.
\end{equation}
Note that we've neglected terms of order $\mathcal O(N^{-1/2})$ in the action (\ref{Vdisc}) and that we've assumed $r_i$ to be the discretization of an at least twice differentiable interpolation of the input measurements.

For strongly varying rain input or high-frequency output data, $\Delta t$ needs to be chosen small. Thus, the harmonic part of the Hamiltonian becomes stiff and, consequently, the time-step of MD needs to be chosen small, which leads to slow convergence.
To solve this problem, we change coordinates, such that the harmonic part of the Hamiltonian becomes (partly) diagonal and the particle interaction moves to the potential.
One option would be to use the Fourier modes of ${\eta}(t)$ as configurational variables. Calculating the potential requires then an inversion of a Toeplitz matrix. This approach is interesting because, in our case, it allows for an approximate analytical integration of high-frequency modes as outlined in Ref.~\cite{doll_1985}.

We choose however another approach, the so-called \emph{staging} as described in Ref.~\cite{tuckerman_1992}, and diagonalize the harmonic part only in between the measurement points.
We first note that the Hamiltonian (\ref{Hamiltonian}) can be seen as the classical Hamiltonian of a polymer chain with harmonic bonds between neighbouring beads, described by the first harmonic term of $V(q,\bt)$, in an external field described by the other non-harmonic terms of $V(q,\bt)$.
Furthermore, the model parameters are viewed as additional degrees of freedom that couple with all other degrees of freedom.
Sampling system realizations, each defined by a set of coordinates $\left\{ q_i\right\}$ and parameters $\bt$, is equivalent to sampling states in the polymer dynamics.

The discretized harmonic part of the action is rewritten as
\begin{eqnarray}
  \sum_{i=2}^{N}
  \frac{T}{2\Delta t}
  (q_i-q_{i-1})^2 \nonumber
  \\ =
  \frac{T}{2}
  \sum_{s=1}^{n}\left\{
    \frac{(q_{(s-1)j+1} - q_{sj+1})^2}{j\Delta t}
    +
    \sum_{k=2}^j
    \frac{k}{(k-1)\Delta t}
    (q_{(s-1)j+k}-q^*_{(s-1)j+k})^2
  \right\}\,,\nonumber
  \\
\end{eqnarray}
with
\begin{equation}
  q^*_{(s-1)j+k}
  =
  \frac{(k-1)q_{(s-1)j+k+1} + q_{(s-1)j+1} }{k}
  \,.
\end{equation}
Following Ref.~\cite{tuckerman_1993} we apply the coordinate transformation
\begin{equation}
  u_{sj+1} = q_{sj+1}\,,\quad
  u_{sj+k} = q_{sj+k} - q^*_{sj+k}\,,
\end{equation}
with its inverse given by
\begin{equation}
  q_{sj+1} = u_{sj+1}\,,\quad
  q_{sj+k} = \sum_{l=k}^{j+1}\frac{k-1}{l-1}u_{sj+l}
  +\frac{j-k+1}{j}u_{sj+1}\,,
\end{equation}
or equivalently by the recursive relation
\begin{equation}
  q_{sj+k} = u_{sj+k} + \frac{k-1}{k} q_{sj+k+1}+ \frac{1}{k}u_{sj+1} \,.
\end{equation}
